# Ten Quick Single-Site Sensitivity Analyses


PEcAn's existing sensitivity analysis tool was built assuming a single-site run.
When turned on in a multi-site settings, it will attempt to run for all sites,
but will then the post-run analysis for each site will all use the same filenames,
so each site overwrites the SA results from the previous one.

We will correct this presently, but meanwhile here's a low-effort workaround:
I'll run the workflow for six different sites into six different output
directories and run the SA separately for each site using only the relevant PFT.

```{r setup}
library(tidyverse)
```


### Site selection

Random for now, choosing five deciduous and five grass sites.
I'd eventually like to use anchor sites for this, but the existing design
points file doesn't provide an immediately obvious way to find the mapping back
from design points to flux measurement points.
TODO: fix that too.


```{r pick-sa-sites}
set.seed(433563456)
sa_sites <- read.csv("site_info.csv") |>
  group_by(site.pft) |>
  sample_n(5)
```


### File manipulation


To turn on sensitivity analysis, we add a `<sensitivity.analysis>` block to the
existing settings file.
N.B. I am _not_ parsing the XML here, just treating whole lines as strings
and inserting the SA block before the final `</pecan>`.

PEcAn will generate one model that holds all parameters at their median,
plus one model per specified quantile _for each input parameter_ defined by the
PFT, so the six quantiles listed here create  a total of 1 + (6 x 22) = 133
simulations at each grass site and 1 + (6 x 28) = 169 simulations at each
deciduous site.

After all simulations are run, the results are analyzed separately for their
effect on each output variable, averaged across 2016-2023.
Adding more output variables does not change the number of model invocations
used in the analysis.


```{r add-sa-block}
template_lines <- readLines("template.xml")
sa_block <- c(
  "<sensitivity.analysis>",
  "  <quantiles>",
  "   <sigma>-3</sigma>",
  "   <sigma>-2</sigma>",
  "  <sigma>-1</sigma>",
  "  <sigma>1</sigma>",
  "  <sigma>2</sigma>",
  "  <sigma>3</sigma>",
  " </quantiles>",
  " <variable>NPP</variable>",
  " <variable>TotSoilCarb</variable>",
  " <variable>AbvGrndWood</variable>",
  " <variable>Qle</variable>",
  " <variable>SoilMoistFrac</variable>",
  " <perpft>TRUE</perpft>",
  " <start.year>2016</start.year>",
  " <end.year>2023</end.year>",
  "</sensitivity.analysis>"
)
cat(template_lines[-length(template_lines)], file = "template_sa.xml", sep = "\n")
cat(sa_block, file = "template_sa.xml", append = TRUE, sep = "\n")
cat(template_lines[length(template_lines)], file = "template_sa.xml", append = TRUE, sep = "\n")
```

Generate a separate site info csv and settings xml for each site to be run.
While we're editing the XML we also reduce the ensemble size to 1 and turn off
retention of raw model output, for a ~70% reduction in output disk space.

Note this requires PEcAn.settings 1.9.1 or later,
to fix a bug that occurred when `site_info.csv` contained only one site
and caused the XML output to wrap the entire list in an extra `<info>` element.


```{r settings-build-fn}
build_site_sa_files <- function(site) {
  site_info <- read.csv("site_info.csv") |>
    filter(id == site)
  write.csv(site_info, paste0("site_info_", site, ".csv"), row.names = FALSE)

  s <- PEcAn.settings::read.settings("template_sa.xml")

  site_out <- paste0("output_", site)
  s$outdir <- site_out
  s$rundir <- file.path(site_out, "run")
  s$modeloutdir <- file.path(site_out, "out")
  s$host$outdir <- s$outdir
  s$host$rundir <- s$rundir
  s$model$delete.raw <- TRUE
  s$pfts <- s$pfts |> keep(\(x)x$name == site_info$site.pft)
  PEcAn.settings::write.settings(
    settings = s,
    outputfile = paste0("template_", site, ".xml"),
    outputdir = "."
  )

  callr::rscript(
    "03_xml_build.R",
    cmdargs = c(
      "--n_ens=1",
      "--n_met=1",
      paste0("--site_file=site_info_", site, ".csv"),
      paste0("--template_file=template_", site, ".xml"),
      paste0("--output_file=settings_", site, ".xml")
    )
  )
}

sa_sites |>
  pull(id) |>
  walk(build_site_sa_files)
```


```{r run-sa}
run_site_sa <- function(site) {
  callr::rscript(
    "04_run_model.R",
    cmdargs = paste0("--settings=settings_", site, ".xml")
  )
}

sa_sites |>
  pull(id) |>
  walk(run_site_sa)
```

Each site's output contains PDFs with figures summarizing the results of the
sensitivity analysis, but also a set of Rdata files with the computed values.

Let's collect the Rdata from all sites into a single dataframe.

```{r collect-sa-data}
read_one_vardecomp <- function(file) {
  sa_res <- PEcAn.utils::load_local(file)$sensitivity.results
  pft_name <- names(sa_res)[[1]]

  sa_res[[pft_name]] |>
    _$variance.decomposition.output |>
    as.data.frame() |>
    rownames_to_column(var = "parameter") |>
    mutate(
      response_var = sub(".*NOENSEMBLEID\\.([a-zA-Z]+)\\.2016.*", "\\1", file),
      pft = pft_name
    )
}

read_site_vardecomps <- function(dir) {
  list.files(dir, pattern = "sensitivity.results.*.Rdata", full.names = TRUE) |>
    map(read_one_vardecomp) |>
    list_rbind() |>
    mutate(site = sub("output_", "", basename(dir)))
}

var_decomps <- sa_sites |>
  mutate(outdir = paste0("output_", id)) |>
  map(read_site_vardecomps) |>
  list_rbind() |>
  mutate(
    cv = coef.vars * 100,
    var = sqrt(variances)
  )
```

Now make a multisite and multi-response-variable version of the
variance decomposition plots that are in each site's PDF output.

```{r var-decomp-summary}
theme_set(theme_bw())
base_plot <- ggplot(var_decomps) +
  aes(
    xmin = 0,
    y = reorder(parameter, var),
    color = site
  ) +
  geom_pointrange(position = position_jitter(height = 0.1)) +
  facet_grid(pft ~ response_var, scales = "free_x") +
  ylab(NULL)

base_plot +
  aes(x = cv, xmax = cv) +
  xlab("CV (%)")

base_plot +
  aes(x = elasticities, xmax = elasticities) +
  xlab("Elasticity")

base_plot +
  aes(x = var, xmax = var) +
  xlab("Variance")
```

Another way of looking at this: Pick the single parameter accounting for the
most variance in each response, ask if this is the same across sites.

```{r}
var_decomps |>
  group_by(pft, site, response_var) |>
  summarize(across(where(is.numeric), \(x)parameter[which.max(x)])) |>
  knitr::kable()
```
