# Initial Evaluation of Sipnet Irrigation Support

Comparing aboveground wood predictions between irrigated and unirrigated
 runs of Sipnet for 2016-2023 at 98 woody perennial sites across CA.


```{r setup}
library(tidyverse)
library(tidyterra)
library(terra)

irri_path <- "_from_cluster/ccmmf_phase_1b_20250314211904_12855/"
noirri_path <- "_from_cluster/ccmmf_phase_1b_98sites_20reps_20250312/"
landtrendr_2016_path <- "data_raw/ca_biomassfiaald_2016_median.tif"
landtrendr_23_path <- "data_raw/ca_composite_2023_median.tif"
ca_outline_path <- "data_raw/ca_outline_shp/"


```

## Read outputs

Defining a function to read files -- note that this is duplicated in 
validate.Rmd, and the version there may have some improvements.
TODO unify these.

```{r fn}
read_ncdir <- function(path, vars = NULL, pattern = "ENS", ...) {
  path |>
    list.files(pattern = "*.nc$", full.names = TRUE, recursive = TRUE) |>
    grep(pattern = pattern, value = TRUE) |>
    data.frame(file = _) |>
    separate_wider_regex(
      cols = file,
      patterns = c(
        ".*/ENS-",
        ens_num = "\\d+",
        "-",
        site = ".*?",
        "/.*"
      ),
      cols_remove = FALSE
    ) |>
    mutate(ens_num = as.numeric(ens_num)) |>
    nest_by(ens_num, site, .key = "path") |>
    mutate(
      contents = map(
        path,
        \(x)PEcAn.utils::read.output(
          ncfiles = x,
          dataframe = TRUE,
          variables = vars,
          print_summary = FALSE,
          verbose = FALSE,
          ...
        )
      )
    ) |>
    unnest(contents)
}
```


```{r read-data}
abvgrdwood_noirri <- read_ncdir(file.path(noirri_path, "out"), vars = "AbvGrndWood", start.year = 2023)
abvgrdwood_irri <- read_ncdir(file.path(irri_path, "output/out"), vars = "AbvGrndWood", start.year = 2023)
site_info <- read.csv(file.path(irri_path, "site_info.csv"))
```


## Plot onto a map

Mapping each site as a colored point, facetting by ensemble member
...Hard to see much from this TBH. Saving this code but marking it eval=FALSE.

```{r points-on-map, eval=FALSE}
ca <- terra::vect(ca_outline_path)

abvgrdwood_noirri |>
  filter(posix == max(posix)) |>
  left_join(site_info, by = c(site = "id")) |>
  vect(crs = "epsg:4326") |>
  ggplot() +
  geom_spatvector(data = ca) +
  geom_spatvector(aes(color = AbvGrndWood)) +
  theme_bw() +
  facet_wrap(~ens_num) +
  ggtitle("Rainfed only")
abvgrdwood_irri |>
  filter(posix == max(posix)) |>
  left_join(site_info, by = c(site = "id")) |>
  vect(crs = "epsg:4326") |>
  ggplot() +
  geom_spatvector(data = ca) +
  geom_spatvector(aes(color = AbvGrndWood)) +
  theme_bw() +
  facet_wrap(~ens_num) +
  ggtitle("Irrigated 1500 mm/yr")
```

## Plot wood stocks

Density distributions of site wood stocks

```{r stock-density}
agb_tlast <- bind_rows(
  rainfed = abvgrdwood_noirri |> filter(posix == max(posix)),
  irrigated = abvgrdwood_irri |> filter(posix == max(posix)),
  .id = "water_source"
)
agb_tlast |>
  ggplot() +
  aes(
    x = AbvGrndWood,
    color = water_source,
    group = paste(water_source, ens_num)
  ) +
  geom_freqpoly(stat = "density") +
  theme_bw() +
  xlab("Modeled aboveground wood in 2023, kg C m-2") +
  ggtitle("Distribution of wood C stock in 98 orchards")
```

## Compare to LandTrender 2023 medians

```{r vs-ltmed}
lt_med <- terra::rast(landtrendr_23_path)
wood_carbon_fraction <- 0.47

point_buf <- site_info |>
  terra::vect(crs = "epsg:4326") |>
  terra::project(lt_med) |>
  terra::buffer(width = 200)
landtrender_2023_agb <- terra::extract(
  x = lt_med,
  y = point_buf,
  fun = mean,
  bind = TRUE
) |>
  as.data.frame() |>
  select(
    site = id,
    AGB_median = ends_with("median")
  ) |>
  mutate(AGB_median = PEcAn.utils::ud_convert(
    AGB_median * wood_carbon_fraction, "Mg ha-1", "kg m-2"))
agb_with_lt <- agb_tlast |>
  left_join(landtrender_2023_agb)
ggplot(agb_with_lt) +
  aes(AGB_median, AbvGrndWood, color = water_source) +
  geom_abline(lty = "dashed") +
  geom_point() +
  geom_smooth(method = "lm") +
  facet_wrap(~ens_num) +
  theme_bw() +
  xlab("LandTrender AGB 2023, kg C m-2") +
  ylab("Sipnet AGB 2023, kg C m-2")

agb_rmse <- agb_with_lt |>
  ungroup() |>
  nest_by(water_source, ens_num) |>
  mutate(
    rmse = sqrt(mean((data$AGB_median - data$AbvGrndWood)^2))
  )
agb_rmse |>
  group_by(water_source) |>
  summarize(mean(rmse), sd(rmse))
```



## Comparing start and end biomass vs start and end landtrender

```{r compare-2016}
lt_med_16 <- terra::rast(landtrendr_2016_path)
landtrender_2016_agb <- terra::extract(
  x = lt_med_16,
  y = point_buf,
  fun = mean,
  bind = TRUE
) |>
  as.data.frame() |>
  select(
    site = id,
    AGB_median = ends_with("median")
  ) |>
  mutate(
    AGB_median = PEcAn.utils::ud_convert(
      AGB_median * wood_carbon_fraction,
      "Mg ha-1",
      "kg m-2"
    )
  )

landtrender_2016_agb |>
  left_join(landtrender_2023_agb, by = "site", suffix = c("_2016", "_2023")) |>
  mutate(diff = AGB_median_2023 - AGB_median_2016) |>
  ggplot() +
  aes(AGB_median_2016, AGB_median_2023, color = diff) +
  geom_point() +
  geom_abline() +
  geom_smooth(method = "lm") +
  theme_bw()

agb_t0 <- bind_rows(
  rainfed = read_ncdir(
      file.path(noirri_path, "out"),
      vars = "AbvGrndWood",
      start.year = 2016,
      end.year = 2016) |>
    filter(posix == min(posix)),
  irrigated = read_ncdir(
      file.path(irri_path, "output/out"),
      vars = "AbvGrndWood",
      start.year = 2016,
      end.year = 2016) |>
    filter(posix == min(posix)),
  .id = "water_source"
)

landtrender_diff <- landtrender_2016_agb |>
  left_join(landtrender_2023_agb, by = "site", suffix = c("_2016", "_2023")) |>
  mutate(AGB_median_diff = AGB_median_2023 - AGB_median_2016)
agb_t0 |>
  left_join(landtrender_diff) |>
  head()
```