# Comparing LandTrendr and CARB allometric orchard wood biomass predictions

```{r setup}
library(tidyverse)
library(terra)
library(tidyterra)
knitr::opts_chunk$set(fig.width = 6, fig.height = 4.5, out.width = "100%")
```

## Define biomass-from-age prediction function

Combines allometries from CARB for trees per acre and carbon per tree according to age.
Applicable to orange, almond, walnut, or pistachio trees that are 0-50 years old.

TODO incorporate uncertainty from DEMETER_FutureModeling.xlsx.

```{r allom-fn}
allom_kgC_m2 <- function(age, species = c("almond", "walnut", "pistachio", "orange")) {
  if (any(age > 50)) stop("this allometry only valid to age 50")
  species <- match.arg(species)

  dens_m <- c(almond = -15.45, walnut = -19.97, pistachio = -24.7, orange = -34.71)
  dens_b <- c(almond = 125.72, walnut = 103.5, pistachio = 184.98, orange = 223.83)
  carbon_m <- c(almond = 12.9823, walnut = 9.3295, pistachio = 1.2879, orange = 2.4435)
  carbon_b <- c(almond = 1.3923, walnut = 1.6121, pistachio = 1.8835, orange = 1.3712)

  trees_acre <- dens_m[species] * log(age) + dens_b[species]
  lbs_tree <- carbon_m[species] * (age^carbon_b[species])

  PEcAn.utils::ud_convert(trees_acre * lbs_tree, "lb acre-1", "kg m-2")
}
```

## Read LandTrendr biomass from satellite timeseries

I'll use 2023 medians from the LandTrendr composite biomass product ([Kennedy et al. 2018](https://doi.org/10.1088/1748-9326/aa9d9e)),
which is primarily Landsat-derived and refined by data fusion with FIA and lidar forest plot data.
The reported values are aboveground _wood_ mass in Mg/ha, so multiply by 0.47 to get carbon.

```{r read_raw}
lt_map <- rast("data_raw/ca_composite_2023_median.tif")

dwr_map <- vect("data_raw/dwr_map/i15_Crop_Mapping_2023_Provisional_20241127.gdb") |> 
  project(lt_map)
```

Exploratory analysis (not shown here) suggests that the 2023 DWR map does not break out citrus by species:
There are zero fields whose main crop is grapefruit (C1), lemon (C2), or orange (C3),
but it does have fields for most non-citrus crops in the C category
including C4 dates, C5 avocadoes, C6 olives, C7 "miscellaneous subtropical fruit".
I'll take this to mean all types of citrus are lumped as class "C" (unspecified citrus and subtropical)
and will use the orange allometry for all fields with `MAIN_CROP=="C"`.

A `YR_PLANTED` of 1984 (the earliest that appears in the dataset) appears to mean "1984 or earlier",
so I'll use fields listed as planted between 1985 and 2022.

```{r extract}
dwr_allom <- dwr_map |>
  select(UniqueID, YR_PLANTED, MAIN_CROP) |>
  filter(
    YR_PLANTED > 1984,
    YR_PLANTED < 2023,
    MAIN_CROP %in% c("C", "D12", "D13", "D14")
  ) |>
  mutate(
    species = case_when(
      MAIN_CROP == "C" ~ "orange",
      MAIN_CROP == "D12" ~ "almond",
      MAIN_CROP == "D13" ~ "walnut",
      MAIN_CROP == "D14" ~ "pistachio"
  )) |>
  group_by(species) |>
  mutate(allom_C = allom_kgC_m2(2023 - YR_PLANTED, species[[1]]))  

pred_C <- extract(lt_map, dwr_allom, fun = mean, bind = TRUE) |>
  as_tibble() |>
  rename(lt_C = ca_composite_2023_median) |>
  mutate(lt_C = PEcAn.utils::ud_convert(0.47 * lt_C, "Mg ha-1", "kg m-2"))

pred_C |>
  group_by(species) |>
  summarize(across(everything(), c(n=length,mean=mean,sd=sd,max=max,min=min)))
```

```{r plot}
theme_set(theme_bw())

#ggplot(pred_C) +
#  aes(lt_C, allom_C) +
#  facet_wrap(~species) +
#  geom_smooth(method = "lm") +
#  geom_abline(lty = "dashed")

#same as above but with hex tiles
ggplot(pred_C) +
  aes(lt_C, allom_C) +
  # geom_point(alpha = 0.3) +
  geom_hex(binwidth = 0.5) +
  scale_fill_gradient(trans = "log10") +
  facet_wrap(~species) +
  geom_smooth(method = "lm") +
  geom_abline(lty = "dashed")

ggplot(pred_C) +
  aes(2023 - YR_PLANTED, lt_C) +
  facet_wrap(~species) +
  # geom_point(alpha = 0.3) +
  geom_hex(binwidth = 1) +
  scale_fill_gradient(trans = "log10") +
  geom_smooth(method = "lm")
```

```{r stats}
pred_C |>
  ungroup() |>
  nest_by(species) |>
  mutate(
    fit = list(lm(allom_C ~ lt_C, data = data)),
    r2 = summary(fit)$r.squared,
    rmse = sqrt(mean(fit$residuals^2)),
    nse = 1 - (
      sum((data$allom_C - data$lt_C)^2) /
        sum((data$allom_C - mean(data$allom_C, na.rm = TRUE))^2)
    ))

# With 1984 included:
## # A tibble: 4 × 6
## # Rowwise:  species
##   species                 data fit         r2  rmse    nse
##   <chr>     <list<tibble[,5]>> <list>   <dbl> <dbl>  <dbl>
## 1 almond          [44,877 × 5] <lm>   0.0132   3.74 -0.728
## 2 orange          [30,967 × 5] <lm>   0.0536   1.37 -1.94 
## 3 pistachio       [10,768 × 5] <lm>   0.149    3.55 -0.197
## 4 walnut          [18,622 × 5] <lm>   0.00543  3.82 -1.30
```
